---
title: "Statistics for Data Science - UC3M Master"
subtitle: "Multivariate Analysis - First Assignment"
author: 
  - "Ricardo Hortelano"
  - "Javier Martinez"
  - "Santiago Raposo"
header-includes:
   - \usepackage{amsmath}
   - \usepackage{bm}
output:
  pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA)
library(plotly)
library(tidyverse)
library(MASS)
library(pracma)
library(corrplot)
library(GGally)

poke <- read_csv("data/pokemon.csv")[,-1]
poke_long <- poke %>%
  dplyr::select(-target) %>%
  pivot_longer(
    everything(),
    names_to = c("fight", ".value"),
    names_sep = "_"
  )
```

# Part 1 - Pre-process the data set for practical analysis.
The only columns with missing values is the `poke1_Type.2` and `poke2_Type.2`. These missing values are not real missing values. 
An empty value in this field means that pokemon has not a second type. Knowing this, we treat that empty value as a new category.

Regarding the rest of the dataset, there are no missing values.

As we have some knowledge of the structure of our dataset, we might wonder if pokemons which are labeled as `Legendary` are outliers in the distribution of their statistics. We can check this using a parallel coordinates plot.
```{r, cache=TRUE, dependson=setup, message = F}
numeric_vars <- poke_long %>% 
  dplyr::select(-number) %>% 
  select_if(~is.numeric(.) || is.logical(.))
# Take a sample of size = 1000 for plots
numeric_sample <- numeric_vars[sample(nrow(numeric_vars)), ][1:1e3,]
ggpairs(numeric_sample, aes(color = Legendary))
```

In order to check if there are any outliers in our variables, we will

# Part 2 - Perform a descriptive analysis of the pre-processed data set and obtain interesting conclusions from the analysis.
[Insert conclusions here]
